### 2022-10-12 AISTATS2023/paper.tex
revised method description

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}
\bibliographystyle{plainnat}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xfrac}
\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{ {./../examples/} }
% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

@@

\begin{abstract}

Accumulated Local Effects (ALE) is a popular approach to explainable AI that quantifies how a feature influences the decisions of a model taking into account feature correlations. However, in case of strong interactions between features, instance-level feature effects may deviate from the average curve provided by ALE. Therefore, it is crucial to quantify this deviation, namely, the uncertainty of the effect. In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and the uncertainty. Furthermore, as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting). The average effect and the uncertainty are then computed from the instances that lie in each bin.  In this work, we formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints. Based on this, we propose a method for automatically finding the optimal intervals, balancing the trade-off between estimation bias and variance~\todo{check if this phrase explains well the concept.} We demonstrate through synthetic and real datasets (a) the advantages of modeling the uncertainty with UALE compared to other methods and (b) the importance of appropriate bin splitting for an accurate approximation of the average effect and uncertainty.
\end{abstract}

@@

Recently, Machine Learning (ML) has been adopted in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason, there is an increased interest in Explainable
AI (XAI) to understand the decision mechanism of complex black-box
models. XAI literature distinguishes between local and global
techniques~\citep{Molnar2020interpretable}. Local methods provide
instance-level explanations, i.e.~explain the prediction for a
specific instance, whereas global methods summarize the entire model
behavior.

Global methods create a global explanation by aggregating the
instance-level explanations into a single interpretable outcome,
usually a number or a plot. When strong interactions between features
exist, the instance-level explanations may significantly deviate from
the aggregated outcome~\citep{herbinger2022repid}, a phenomenon called
\emph{aggregation bias}~\citep{mehrabi2021survey}. Therefore, global
methods should quantify not only the aggregated (averaged) effect, but
also the level of heterogeneity of instance-level explanations. The
same holds for Feature Effect (FE)~\citep{Gromping2020MAEP}, a class
of global XAI methods that isolate the impact of a single feature on
the output.\footnote{FE methods also isolate the combined effect of a
  pair of features to the output. Combinations of more than two
  features are uncommon, since they are difficult to estimate and
  visualize.} For example, a feature with zero average effect may
indicate no effect on the output or a highly positive effect in some
cases and a highly negative effect in others. The ambiguity about the
correct interpretation can be resolved, if the FE method informs about
the heterogeneity of instance-level explanations, a quantity we refer
to as the \emph{uncertainty of the effect}.

There are three widely-used FE methods: \emph{Partial Dependence
  Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
(ALE)\citep{apley2020visualizing}. ALE has been established as the
state-of-the-art for measuring the average effect, since PDP and MP
have been criticized~\citep{Gromping2020MAEP} to compute erroneous
effects when the input features are (highly) correlated. However, ALE
faces two limitations, the first concerns ALE definition and the
second ALE approximation. Regarding the definition, ALE does not
inform the user about the uncertainty. In contrast, in PDP the
heterogeneity of instance-level effects can be identified by exploring
the Individual Conditional Expectations
(ICE)~\citep{goldstein2015peeking}. Regarding the approximation, ALE
requires partitioning the feature domain into non-overlapping
intervals (bin-splitting) and estimate the effect inside each interval
(bin-effect) from the training set instances that fall inside. Finding
an appropriate sequence of bins is of particular importance, since
ALE's interpretation is meaningful only inside each
bin~\citep{molnar2022}. However, this crucial step has not raised the
appropriate attention. With the current
approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
split \emph{blindly}, i.e.~without considering the characteristics of
the instance-level effects, in \(K\) equisized bins.

In this paper, we present Uncertainty-aware Accumulated Local Effects
(UALE), a method that visualizes in a single plot both the average
feature effect and the uncertainty, i.e.~the level of heterogeneity of
instance-level effects. The average effect is defined as in ALE and
the uncertainty inherits ALE's advantages, i.e.~takes into account the
correlations between features. We also propose an accurate UALE
approximation. We formally prove that for an unbiased estimation of
the uncertainty, the partitioning of the feature domain must respect
specific conditions. Using that, we set-up an optimization problem
where we search for a partitioning, a sequence of \emph{variable-size}
intervals, (a) with enough instances (accurate estimation) and (b)
that respect the above conditions (unbiased estimation). Finally, we
propose a method for automatically finding the optimal
partitioning. We apply our method in multiple synthetic and real world
datasets for evaluating its performance. The contributions of this
paper are:

\begin{itemize}
\item UALE, a feature effect method that quantifies both the average
  effect and the uncertainty, i.e.,~the heterogeneity of instance level
  effects.
\item A formal proof of the conditions the feature domain partitioning
  must fulfill for an unbiased approximation of the uncertainty.
\item An accurate UALE approximation by automatically partitioning
  into variable-size intervals that respect the above conditions.
\end{itemize}

The implementation of our method and the code for
reproducing all the experiments is provided in the submission and will
become publicly available upon acceptance.

@@

We here review the existing methods for modeling the average effect
and the heterogeneity of instance-level effects and we focus on ALE
definition~\citep{apley2020visualizing} and
approximation~\citep{apley2020visualizing, gkolemis22}.

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), to simple variables with plain lowercase \( x \)
and to vectors with bold \( \xb \). Often, we partition the input
vector \(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we it refers the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). When we
refer to an instance of the training set, we use
\(\xb^i= (x_s^i, \xc^i) \).

% The black-box function is
% \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
% feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
% indicating the particular method in use.\footnote{An extensive list of
%   all symbols used in the paper is provided at the Appendix.}

\subsection{Feature Effect Methods}
\label{sec:feat-effect-meth}

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th feature as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect at \(x_s = z\) as
\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). The local
explanations are then weighted by the conditional distribution
\(p(\xc|x_s=z)\) to compute the averaged effect \(\mu(z)\). ALE is the
accumulation of the averaged local effects:

\begin{equation}
  \label{eq:ALE}
  f^{\mathtt{ALE}}(x_s) = \int_{x_{s,min}}^{x_s} \underbrace{\mathbb{E}_{\Xcb|X_s=z}\left [f^s (z, \Xcb)\right ]}_{\mu(z)} \partial z
\end{equation}

Eq.~(\ref{eq:ALE}) has important advantages compared to PDP and MP in
cases of correlated features. In these cases, PDP integrates over
unrealistic instances due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
resolves both issues.

@@

FE methods show what is the average (global) effect on the output, if
the value of a specific feature is increased/decreased. For completing
the explanation about a feature, we need to also know to what extent
the local effects deviate from the global explanation, i.e.~to
quantify the uncertainty of the average effect.

ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
are plot on top-of PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the curves are
homogeneous and to what extent they deviate from the PDP. Some methods
try to automate the aforementioned visual exploration, by grouping
(d-)ICE plots into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent} inform indirectly about
heterogeneous effects, quantifying interactions between the input
features. A strong interaction index is an indicator for the existence
of heterogeneous effects.

The aforementioned approaches are under two limitations. They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features\citep{baniecki2021fooling}, as we will
show in-depth in Section~\ref{sec:simulation-examples-1}. To the best
of our knowledge, no work exist so far that quantifies the
heterogeneous effects based on the formulation of ALE.

@@

In ML scenarios, ALE is estimated from the instances of the training
set.~\citep{apley2020visualizing} proposed dividing the feature domain
in \(K\) equisized bins and estimating the local effects in each bin
by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instances of the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
(DALE) that uses the partial derivatives:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution when
bins become wide and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the bin
density for splitting the feature axis into \(K\) equisized bins, an
approach that is error prone. Setting \(K\) to a small value may hide
fine-grain effects due to large bins while setting \(K\) to a high
value leads to poor bin-effect estimations due to few samples per
bin. In general, the user may receive contradictory explanations for
different \(K\) without a clue for which one to trust.

@@

UALE extends ALE for quantifying the level of heterogeneous effects
(uncertainty) and automates the bin-splitting. At
Section~\ref{sec:UALE-definition-1} we define UALE, at
Section~\ref{sec:UALE-definition-2} we reformulate it using an
interval-based formulation, at Section~\ref{sec:UALE-approximation} we
provide UALE approximation and at Section~\ref{sec:bin-spliting} we
define and solve the problem of optimal bin-splitting.

\subsection{UALE: ALE With Uncertainty Quantification}
\label{sec:UALE-definition-1}

UALE extends ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it with the standard deviation of the local explanations
\(\mathcal{H}(z) = \sigma(z)\), where:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(z) = \mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
UALE formulates the effect at a specific point \(x_s\) with a tuple
that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes it with a continuous curve and
a confidence region
\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

\subsection{Interval-Based Formulation and Approximation}
\label{sec:UALE-definition-2}

In real ML scenarios, all estimations are based on the instances of
the training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\)
at the granularity of a point, because the probability of observing a
sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We define the regional-effect \(\mu(z_1, z_2)\) and the
regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as
the summary statistics that characterize the effect and the
uncertainty inside the interval \([z_1, z_2)\):

@@

%
Intuitively, the regional-effect and the regional-uncertainty quantify
the expected average effect and the expected uncertainty if we
randomly draw a point \(z^*\) given a uniform distribution
\(z^* \sim \mathcal{U}(z_1, z_2)\). If we also define as
\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
domain of the \(s\)-th feature into \(K\) variable-size intervals,
i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can redefine UALE
using an interval-based formulation
\(\tilde{f}^{\mathtt{UALE}}_{\mathcal{Z}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s)
\pm \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s)\), where:

\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%

\subsection{Interval-Based Approximation}
\label{sec:UALE-approximation}

For approximating the mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) we use the set \(\mathcal{S}_k\)
of dataset instances with the \(s\)-th feature lying inside the
interval
\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):

>>>>

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}
\bibliographystyle{plainnat}
\usepackage{amsmath}
\usepackage{amssymb}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

\usepackage{xfrac}
\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{ {./../examples/} }
% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

@@

\begin{abstract}
Accumulated Local Effects (ALE) is a popular explainable AI method that quantifies how a feature influences the decisions of a model, handling well feature correlations. In case of complex interactions between features, instance-level feature effects may deviate from the ALE curve. It is therefore crucial to quantify this deviation, namely, the uncertainty of the effect. In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and its uncertainty. We show that UALE quantifies uncertainty effectively, even in case of correlated features. We also note that as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting). The average effect and the uncertainty are computed from the instances that lie in each bin. We formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints. Based on this, we propose a method to determine the optimal intervals, \todo{balancing the estimation bias and variance.} We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods and (b) the effectiveness of UALE's appropriate bin splitting for a \todo{good} approximating of the average effect and its uncertainty.
\end{abstract}

@@

Recently, Machine Learning (ML) has been widely adopted in mission critical domains,
such as healthcare and finance. In such high-stakes areas,  it is important not only to provide accurate predictions but also meaningful explanations. For this reason, there is an increased interest in Explainable
AI (XAI). XAI literature distinguishes between local and global
methods~\citep{Molnar2020interpretable}. Local methods provide
instance-level explanations, i.e., explain the prediction for a
specific instance, whereas global methods summarize the entire model
behavior. Global methods create a global explanation by aggregating the
instance-level explanations into a single interpretable outcome,
usually a number or a plot.

A popular class of global XAI are feature effect (FE)
methods~\citep{Gromping2020MAEP} that quantify the average (over all
instances) effect of a single feature on the output\footnote{FE
  methods also isolate the combined effect of a pair of features to
  the output. Combinations of more than two features are uncommon,
  since they are difficult to estimate and visualize.}. There are
three widely-used FE methods: \emph{Partial Dependence Plots}
(PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
(MP)\citep{apley2020visualizing} and \emph{Accumulated Local Effects}
(ALE)\citep{apley2020visualizing}. ALE is established as the
state-of-the-art in quantifying the average effect, since PDP and MP
have been criticized~\citep{Gromping2020MAEP} of providing misleading explanations in case of  correlated input features.

When complex interactions between features exist, the instance-level
(local) feature effects may exhibit significant deviation, aka heterogeneity, from the aggregated (global)
explanation, a phenomenon known as aggregation
  bias~\citep{mehrabi2021survey}. We call this deviation \emph{the uncertainty of the global effect}. Aggregation bias leads to
ambiguities; for example, a feature with zero average effect may
indicate (a) no effect on the output or (b) an effect that is highly
positive for some instances and highly negative for some others. As a
result, there is a need for FE methods to quantify the deviation (uncertainty) of
instance-level explanations, in addition to the average effect. In the
case of PDP, such deviation is visualized via the Individual
Conditional Explanations (ICE)~\citep{goldstein2015peeking}. ICE
plots, however, have the same limitations as PDP in case of correlated
features. No method to quantify the deviation of instance-level effects has been proposed for ALE.

% Such ambiguities could be resolved if the heterogeneity of instance-level feature effects is provided in addition to the average effect. 
% In this paper, we measure the heterogeneity as the standard deviation of the instance-level effects and we refer to it using as the \emph{uncertainty of the effect}.

%Therefore, global methods should quantify not only the aggregated (averaged) effect, but also the level of heterogeneity of instance-level explanations.

%There are three widely-used FE methods: \emph{Partial Dependence Plots} (PDP)\citep{friedman2001greedy}, \emph{Marginal Plots}
%(MP)\citep{apley2020visualizing} and \emph{Aggregated Local Effects}
%(ALE)\citep{apley2020visualizing}. ALE has been established as the
%state-of-the-art for measuring the average effect, since PDP and MP
%have been criticized~\citep{Gromping2020MAEP} to compute erroneous
%effects when the input features are (highly) correlated.

In this work, we propose UALE, a global feature-effect method that
extends ALE for modeling the uncertainty of the feature effect.
Our method handles well cases with correlated features. 
As in ALE, we provide an interval-based formulation of UALE. 
This formulation partitions the feature domain into non-overlapping 
intervals (bin-splitting) and defines the average effect (bin-effect) and its uncertainty (bin-uncertainty). We exploit this formulation to provide an UALE approximation. 
As we formally prove, a fixed equi-width partitioning that does not consider
the characteristics of the instance-level effects, which is the case in ALE, 
may lead to an erroneous (biased) estimation of the uncertainty. 
Therefore, we propose a variable-width partitioning that leads to an unbiased approximation.

% %However, ALE faces two limitations, the first concerns ALE definition and the second ALE approximation. Regarding the definition, ALE does not inform the user about the uncertainty. 
% \color{blue}\todo{This should go to related work}In contrast, in PDP the
% heterogeneity of instance-level effects can be identified by exploring
% the Individual Conditional Expectations
% (ICE)~\citep{goldstein2015peeking}. \color{black}
% %Regarding the approximation, ALE requires partitioning the feature domain into non-overlapping intervals (bin-splitting) and estimate the effect inside each interval (bin-effect) from the training set instances that fall inside. 
% \color{blue}\todo{to be moved in related work}
% Finding
% an appropriate sequence of bins is of particular importance, since
% ALE's interpretation is meaningful only inside each
% bin~\citep{molnar2022}. However, this crucial step has not raised the
% appropriate attention. With the current
% approaches~\citep{apley2020visualizing, gkolemis22}, the domain is
% split \emph{blindly}, i.e.~without considering the characteristics of
% the instance-level effects, in \(K\) equisized bins.\color{black}

%In this paper, we present Uncertainty-aware Accumulated Local Effects(UALE), a method that visualizes in a single plot both the av erage feature effect and the uncertainty, i.e.~the level of heterogeneity of instance-level effects. The average effect is defined as in ALE and the uncertainty inherits ALE's advantages, i.e.~takes into account the correlations between features. We also propose an accurate UALE approximation. We formally prove that for an unbiased estimation of the uncertainty, the partitioning of the feature domain must respect specific conditions. Using that, we set-up an optimization problem where we search for a partitioning, a sequence of \emph{variable-size}intervals, (a) with enough instances (accurate estimation) and (b) that respect the above conditions (unbiased estimation). Finally, we propose a method for automatically finding the optimal partitioning. We apply our method in multiple synthetic and real world datasets for evaluating its performance. 
The contributions of this
paper are:
\begin{itemize}
\item A global feature effect method (UALE) that quantifies both the
  average effect and its uncertainty, i.e.,~the deviation of
  instance-level effects.
\item An unbiased approximation of the uncertainty through a dynamic
  variable-size bin partitioning of the feature domain.
%A formal proof of the conditions the feature domain partitioning must fulfill for an unbiased approximation of the uncertainty.
\item The evaluation of UALE on both synthetic and real datasets.
%An accurate UALE approximation by automatically partitioning into variable-size intervals that respect the above conditions.
\end{itemize}

The implementation of our method and the code for reproducing all the
experiments is provided along with the manuscript and will become
publicly available upon acceptance.

@@

In the rest of the paper, we use following notation. Let \(\mathcal{X} \in \mathbb{R}^d\) be the \(d\)-dimensional feature
space, \(\mathcal{Y} \in \mathbb{R}\) the target space and
\(f(\cdot) : \mathcal{X} \rightarrow \mathcal{Y}\) the black-box
function. We use index \(s \in \{1, \ldots, d\}\) for the feature of
interest and \(c = \{1, \ldots, d\} - s\) for the set with all other
indexes. For convenience, we denote the feature vector
\(\xb = (x_1, \cdots , x_s, \cdots, x_D)\) with \((x_s, \xc)\) and the
corresponding random variables
\(X = (X_1, \cdots , X_s, \cdots, X_D)\) with \((X_s, X_c)\). The
training set is \(\mathcal{D} = \{(\xb^i, y^i)\}_{i=1}^N\) sampled
i.i.d. from the distribution \(\mathbb{P}_{X,Y}\). Finally, we use
\(f^{\mathtt{<method>}}(x_s)\) for denoting the \(s\)-th FE, where
\(\mathtt{<method>}\) indicates the particular method in use, for
example ALE. An extensive list with all symbols used in this paper is
provided at Appendix~\ref{sec:list-with-symbols}.

\subsection{Feature Effect Methods}
\label{sec:feat-effect-meth}
The three well-known feature effect methods are: \emph{Partial
  Dependence Plots} (PDP), \emph{Marginal Plots} (MP) and
\emph{Accumulated Local Effects} (ALE).  PDP formulates the global FE
as an expectation over the distribution of \(X_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) = \mathbb{E}_{X_c}[f(x_s,X_c)]\), whereas MP
as an expectation over the distribution of \(X_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{X_c|x_s}[f(x_s, X_c)]\). Both
methods suffer from misestimations in case of correlated features. PDP
integrates over unrealistic instances and MP computes aggregated
effects, i.e., imputes the combined effect of sets of features to a
single feature~\citep{apley2020visualizing}.

ALE overcomes these problems. Specifically, ALE defines the local
effect of the \(s\)-th feature at a specific point \((z, \xc)\) of the
input space \(\mathcal{X}\) with the partial derivative
\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
local explanations at \(z\) are then weighted by the conditional
distribution \(p(\xc|z)\) and are averaged, to produce the averaged
effect \(\mu(z)\). ALE plot is the accumulation of the averaged local
effects:

\begin{equation}
  \label{eq:ALE}
  f^{\mathtt{ALE}}(x_s) = \int_{x_{s,min}}^{x_s} \underbrace{\mathbb{E}_{X_c|X_s=z}\left [f^s (z, X_c)\right ]}_{\mu(z)} \partial z
\end{equation}
%
where \(x_{s,min}\) is the minimum value of the \(s\)-th feature.

@@

The global effect is computed as an expectation over local
(instance-level) effects. In addition to this global effect, it is
important to know to what extent the local effects deviate from the
global explanation.

ICE plots and similar methods (e.g., d-ICE
plots~\citep{goldstein2015peeking}) provide a set of curves
illustrated on top-of PDP. Each curve corresponds to one instance of
the dataset, \(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\). The user
then visually observes if the curves are homogeneous, i.e., all
instances have similar effect plots, and to what extent they deviate
from the PDP. There are methods try to automate the aforementioned
visual exploration, by grouping ICE plots into
clusters~\citep{molnar2020model, herbinger2022repid,
  britton2019vine}. Unfortunately, these methods are subject to the
failure modes of PDPs in cases of correlated
features~\citep{baniecki2021fooling}, as we also confirm in our
experimental analysis (c.f., Section~\ref{sec:simulation-examples-1}).

Other approaches, like H-Statistic \citep{friedman2008predictive},
Greenwell's interaction index \citep{greenwell2018simple} or SHAP
interaction values \citep{lundberg2018consistent} quantify
interactions between the input features. A strong interaction is
  an implicit indication of the existence of heterogeneous effects. These
  methods, however, do not quantify directly the level of heterogeneity.

To the best of our knowledge, no work exist so far that quantifies the
heterogeneous effects (uncertainty) based on the formulation of ALE.

@@

ALE is estimated from the available dataset instances. \citep{apley2020visualizing} proposed to divide the feature
domain in \(K\) bins of equal size and to estimate the local effects
in each bin by evaluating the model \(f\) at the bin limits:

\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
%
where \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instances of the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). \citep{gkolemis22} proposed the Differential ALE (DALE) that
computes the local effects on the training instances using
auto-differentiation:
\begin{equation}  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has significant computational advantages and allows the
recomputation of the accumulated effect with different bin-splitting
with near-zero additional computational cost. Moreover, by avoiding
the use of artificial samples at the bin limits DALE allows the use of
wider bins without out-of-distribution sampling. \todo{Maybe connect with our work}. In both cases, the
approximations partition the feature domain in \(K\) equally-sized
bins, without considering the underlying local effects.

% as
% input and perform an equally-sized bin splitting of the attribute
% domain. The choice of \(K\) is therefore critical and introduces
% significant limitations. Specifically, setting \(K\) to a small value
% may hide fine-grain effects due to large bins while setting \(K\) to a
% high value leads to poor bin-effect estimations due to few samples per
% bin. In general, the user may receive contradictory explanations for
% different \(K\) values without a clue for which one to trust.

@@

In this Section~\ref{sec:UALE-definition-1}, we define UALE. Then, we
provide an interval-based formulation that partitions the feature
domain into non-overlapping intervals (bin-splitting) and defines the
bin-effect and the bin-uncertainty
(Section~\ref{sec:UALE-definition-2}). Then, we present the UALE
approximation (Section~\ref{sec:UALE-approximation}), formally proving
the bin-splitting requirements for an unbiased estimation of the
uncertainty. Based on this, we present an optimization method to
determine an appropriate variable-width partitioning
(Section~\ref{sec:bin-spliting}).

\subsection{UALE Definition And Interval-Based Formulation}
\label{sec:UALE-definition-1}

We quantify the uncertainty of the local effects at \(x_s=z\) with the
standard deviation of the local explanations, \(\sigma(z)\), where:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(z) = \mathbb{E}_{X_c|X_s=z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
\end{equation}
\noindent
The uncertainty emerges from (a) the feature correlations and (b) the
implicit feature interactions of the black-box function. We also
define the accumulated uncertainty at \(x_s\), as the accumulation of
the standard deviation of the local effects:

\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent

UALE defines the effect as a pair that consists of the average effect
and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we
use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}),
and the accumulated uncertainty, as defined in
Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates
an example \todo{Describe after putting the figure}.

\begin{figure}
  \centering
  \caption{UALE-concept-figure}
  \label{fig:UALE-figure}
\end{figure}

For the interval-based formulation, we define the bin-effect
\(\mu(z_1, z_2)\) and the corresponding bin-uncertainty
\(\sigma(z_1, z_2)\) as:

@@

%
Intuitively, if we randomly draw a point \(z^*\) given a uniform
distribution \(z^* \sim \mathcal{U}(z_1, z_2)\), the bin-effect
(Eq.~(\ref{eq:mu_bin})) and the bin-uncertainty
(Eq.~(\ref{eq:var_bin})) are the expected average effect and the
expected uncertainty, respectively. Also, if we define \(\mathcal{Z}\)
to be the sequence of \(K+1\) points that partition the domain of the
\(s\)-th feature into \(K\) variable-size intervals, i.e.
\(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can reformulate UALE to:
\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
where \(k_x\) is the index of the bin such that
\( z_{k_x - 1} \leq x_s < z_{k_x}\).

\subsection{UALE Interval-Based Approximation}
~\label{sec:UALE-approximation}

For approximating the bin-effect and the bin-uncertainty, we use the
set \(\mathcal{S}_k\) of dataset instances with the \(s\)-th feature
lying inside the \(k\)-th bin, i.e.,
\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \}
\). The bin-effect is approximated with,

\begin{equation}
  \label{eq:mu_bin_approx}
  \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
  \sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i)
  \right ]
\end{equation}

%
which, as shown by\citep{gkolemis22}, is an unbiased estimator of
Eq.~(\ref{eq:mu_bin}) under the assumption that the points are
uniformly distributed inside the intervals. The approximation for the
uncertainty (Eq.~(\ref{eq:var_bin})) from the available instances is
defined as follows:
